Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-07-07 curvature.tex
Tidying up quotes (cc @cwhidden)

It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
We will call these "graph effects"---one can't easily move from one tree to another due to the structure of the graph.
Researchers have identified theoretical SPR graph effects in MCMC previously~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
In particular, the simple case of sampling the uniform topology distribution without continuous parameters via SPR operations is known to mix quickly~\cite{spade2014note}.
However, despite indications of graph effects on real data~\cite{beiko2006searching, lakner2008efficiency} the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations has made it difficult to study graph effects.
By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by "bottlenecks" of specific classes of SPR moves.

@@

In response, we considered properties of the graph analytically.
By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.
In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which modify large subtrees may have greater graph effects than those that move small subtrees.
Pairs of trees separated by such moves correspond to the "bottlenecks" identified in our simulation results.
These pairs occur infrequently in these well-connected graphs, thus allowing the Markov chain to mix rapidly.
However, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\cite{Whidden2015-yi}. 
In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.

@@

\section{Rooted SPR Neighborhoods}
\label{sec:neighborhoods}
Having made the connection between curvature values and access times on rSPR graphs, we now consider curvature theoretically.
We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call "squares" and "triangles" in the rSPR graph.
Many of our results in this section follow from a characterization of the change in degree and distribution of permissible rSPR moves after an rSPR move is applied.

>>>>

It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
We will call these ``graph effects"---one can't easily move from one tree to another due to the structure of the graph.
Researchers have identified theoretical SPR graph effects in MCMC previously~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
In particular, the simple case of sampling the uniform topology distribution without continuous parameters via SPR operations is known to mix quickly~\cite{spade2014note}.
However, despite indications of graph effects on real data~\cite{beiko2006searching, lakner2008efficiency} the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations has made it difficult to study graph effects.
By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by ``bottlenecks" of specific classes of SPR moves.

@@

In response, we considered properties of the graph analytically.
By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.
In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which modify large subtrees may have greater graph effects than those that move small subtrees.
Pairs of trees separated by such moves correspond to the ``bottlenecks" identified in our simulation results.
These pairs occur infrequently in these well-connected graphs, thus allowing the Markov chain to mix rapidly.
However, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\cite{Whidden2015-yi}.
In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.

@@

\section{Rooted SPR Neighborhoods}
\label{sec:neighborhoods}
Having made the connection between curvature values and access times on rSPR graphs, we now consider curvature theoretically.
We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call ``squares" and ``triangles" in the rSPR graph.
Many of our results in this section follow from a characterization of the change in degree and distribution of permissible rSPR moves after an rSPR move is applied.
